Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-07-07 paper/paper.tex
restructure section 6.2

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10003752.10003753.10010622</concept_id>
<concept_desc>Theory of computation~Abstract machines</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
<concept_desc>Software and its engineering~Functional languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation~Abstract machines}
\ccsdesc[500]{Software and its engineering~Functional languages}
%% End of generated code

@@

\begin{lstlisting}
  def ndcontinue(nds: NDState, seen: Set[State]): Set[State] = {
    val NDState(ae, env, bstore, konts, time, ndk) = nds
    ndk match {
      case Nil => seen
      case NDCont(Nil, _, _, _, _)::ndk =>
        drive_step(NDState(ae, env, bstore, konts, time, ndk), seen)
      case NDCont(cls, argvs, bstore, time, frames)::ndk =>
        val Clos(Lam(v, body), c_env) = cls.head
        val baddr = allocBind(v, time);  val new_env = c_env + (v -> baddr)
        val new_store = bstore.update(baddr, argvs)
        val new_ndk = NDCont(cls.tail, argvs, bstore, time, frames)::ndk
        drive_step(NDState(body, new_env, new_store, frames, time, new_ndk, seen))
    } }
\end{lstlisting}

Function @ndcontinue@ is identical to the dispatching logic in @continue@ when @konts@ is empty.
When @ndk@ is empty, the current set of explored states @seen@ is returned, after which the analysis terminates.
Otherwise, a new @NDState@ is constructed depending on the status of the closures set.
% first call to drive_step can be replaced with continue. ae is atomic, konts is empty.

@@

\begin{lstlisting}
  def makeMCont(cls: List[Clos]): MCont =
    if (cls.isEmpty) mcontinue else { (config, seen) =>
      val Clos(Lam(v, body), c_env) = cls.head
      val baddr = allocBind(v, new_time);  val new_env = c_env + (v -> baddr)
      val new_store = store.update(baddr, argvs)
      aeval(Config(body, new_env, new_store, new_time), seen, new_cont, makeMCont(cls.tail)) }
  val new_mcont: MCont = makeMCont(closures.tail)
\end{lstlisting}

@@

\begin{itemize}
  \item The components of the configuration are lifted as arguments of the evaluation
    function.
  \item The refunctionalized AAM collects all possible configurations, but our final target is an
    abstract definitional interpreter that cares about final evaluation results.
    As such, starting from this section, our abstract semantic artifact returns a set of final
    values instead of collected configurations.
  \item The configurations still play an important role in caching, where a cache
    is a mapping from configurations to sets of final values. The caching algorithm
    ensures termination of the analysis, as well as performs a sound over-approximation
    of all possible values and reachable paths.
  \item The function @aeval@ is simplified by introducing and using a nondeterministic
    operator @nd@ which makes @aeval@ only uses one layer of continuation.
\end{itemize}

@@

\subsubsection{Values}
To properly represent final values, we introduce the case class @VS@ which
contains a set of storable values, a timestamp, and a store.

@@

\subsubsection{Cache-Passing Style}
% why cache? problem of set of states.
Recall that the latent assumption of the work-list algorithm in classical AAM is that
if we have seen a state @s@, it means we also have seen all the successors of state @s@.
The caching algorithm we adopt here replays the same assumption, but in a big-step
manner: if we have seen the configuration @c@, then it means we also have seen the
values that are evaluated from the configuration @c@.
Here we will apply the fixed-point caching algorithm as described from
\citeauthor{darais2017abstracting}'s ADI paper \cite{darais2017abstracting}.
The case class @Cache@ and its operations are defined as follow:

@@

We can now transform our abstract interpreter into cache-passing style. We begin by defining the
return type @Ans@ as a case class which contains a @VS@ set and a cache. @Ans@ also implements
its own accumulating operation @++@ which defines as an union of @Set[VS]@s and a join of caches.
The join operation of caches is defined as joining its two store components.

@@

As previously mentioned, the components of configurations are lifted as arguments to
@aeval@. Given the cache and the simplified one-layer continuation (as we will see later),
the shape of function @aeval@ now looks as follows:

\begin{lstlisting}
  type Cont = Ans => Ans
  def aeval(e: Expr, env: Env, store: BStore, time: Time, cache: Cache, continue: Cont): Ans = {
    val config = Config(e, env, store, time);  val new_time = config.tick
    if (cache.outContains(config)) continue(Ans(cache.outGet(config), cache))
    else {
      val new_cache = cache.outUpdate(config, cache.inGet(config))
      e match { ... } } }
\end{lstlisting}

@@

Given a collection of elements of type @T@, an initial accumulator of
type @Ans@, and a function @k@ that works on elements of type @T@, @nd@
chooses an element in
the collection and applies @k@ to it iteratively, and accumulates the values that
returned from @k@ when recurring on the rest of the collection. Eventually, @nd@
returns the accumulated value @acc@ when there is no element in the collection.
The latest accumulated cache @acc.cache@ is exposed to @k@ when applied.

@@

Now, as we can see, by using the @nd@ operator which abstracts over the meta-continuation,
the evaluation function can be simplified to a single-layer continuation-passing style.
The use of two layers of continuations are unchained to two functions,
in that each of them has one continuation.
The only continuation of @aeval@ plays the same role as the one in a
concrete CPS interpreter.
Furthermore, we add a cache along with the abstract interpreter and the @nd@ operator.

>>>>

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10003752.10003753.10010622</concept_id>
<concept_desc>Theory of computation~Abstract machines</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
<concept_desc>Software and its engineering~Functional languages</concept_desc>
<concept_significance>300</concept_significance>
</concept>
<concept>
<concept_id>10011007.10011006.10011041.10010943</concept_id>
<concept_desc>Software and its engineering~Interpreters</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation~Abstract machines}
\ccsdesc[300]{Software and its engineering~Functional languages}
\ccsdesc[300]{Software and its engineering~Interpreters}
%% End of generated code

@@

\begin{lstlisting}
  def ndcontinue(nds: NDState, seen: Set[State]): Set[State] = {
    val NDState(ae, env, bstore, konts, time, ndk) = nds
    ndk match {
      case Nil => seen
      case NDCont(Nil, _, _, _, _)::ndk =>
        /* drive_step() is equivalently replaced by ndcontinue() */
        ndcontinue(NDState(ae, env, bstore, konts, time, ndk), seen)
      case NDCont(cls, argvs, bstore, time, frames)::ndk =>
        val Clos(Lam(v, body), c_env) = cls.head;  val baddr = allocBind(v, time)
        val new_env = c_env + (v -> baddr);  val new_store = bstore.update(baddr, argvs)
        val new_ndk = NDCont(cls.tail, argvs, bstore, time, frames)::ndk
        drive_step(NDState(body, new_env, new_store, frames, time, new_ndk, seen))
    } }
\end{lstlisting}

Function @ndcontinue@ is identical to the dispatching logic in @continue@ when @konts@ is empty,
except that the first application of @drive_step@ is reduced to @ndcontinue@ (we can safely do this because 
@ae@ is atomic and @knots@ is empty).
When @ndk@ is empty, the current set of explored states @seen@ is returned, after which the analysis terminates.
Otherwise, a new @NDState@ is constructed depending on the status of the closures set.

@@

\begin{lstlisting}
  def makeMCont(cls: List[Clos]): MCont =
    if (cls.isEmpty) mcontinue else { (config, seen) =>
      val Clos(Lam(v, body), c_env) = cls.head;  val baddr = allocBind(v, new_time)
      val new_env = c_env + (v -> baddr);  val new_store = store.update(baddr, argvs)
      aeval(Config(body, new_env, new_store, new_time), seen, new_cont, makeMCont(cls.tail)) }
  val new_mcont: MCont = makeMCont(closures.tail)
\end{lstlisting}

@@

\begin{itemize}
  \item The function @aeval@ is simplified by introducing and using a nondeterministic
    operator @nd@ which makes @aeval@ only uses one layer of continuation.
  \item The components of the configuration are lifted as arguments of the evaluation
    function.
  \item The refunctionalized AAM collects all possible configurations, but our final target is an
    abstract definitional interpreter that cares about final evaluation results.
    As such, starting from this section, our abstract semantic artifact returns a set of final
    values instead of collected configurations.
  \item The configurations still play an important role in caching, where a cache
    is a mapping from configurations to sets of final values. The caching algorithm
    ensures termination of the analysis, as well as performs a sound over-approximation
    of all possible values and reachable paths.
\end{itemize}

@@

\begin{lstlisting}
  type Ans = Set[Config]
  def nd[T](ts: Iterable[T], acc: Ans, k: (T, Ans) => Ans): Ans =
    if (ts.isEmpty) acc else nd(ts.tail, acc ++ k(ts.head, acc), k)
\end{lstlisting}

As a matter of convenience, we name the return type of the abstract interpreter to be @Ans@,
which is @Set[Config]@ for the moment.
Given a collection of elements of type @T@, an initial accumulator of type @Ans@, and a function
@k@ that works on elements of type @T@, @nd@ chooses an element in the collection and applies @k@
to it iteratively, and accumulates the values that returned from @k@ when recurring on the rest 
of the collection. Eventually, @nd@ returns the accumulated value @acc@ when there is no element
in the collection. The latest accumulated value @acc@ is exposed to @k@ when applied.

Now we may refactor the function @aeval@ to one layer of continuation by using 
the @nd@ operator.
The use of two layers of continuations will be unchained to two functions, in that each of them 
has one continuation.

\begin{lstlisting}
  type Cont = (Config, Ans) => Ans
  def aeval(config: Config, seen: Ans, continue: Cont): Ans = { ...  e match {
    case Let(x, App(f, ae), e) =>
      val closures = atomicEval(f, env, store).toList.asInstanceOf[List[Clos]]
      nd[Storable](closures, new_seen, { case (Clos(Lam(v, body), c_env), seen) =>
        val baddr = allocBind(v, new_time);      val new_env = c_env + (v -> baddr)
        val argvs = atomicEval(ae, env, store);  val new_store = store.update(baddr, argvs)
        aeval(Config(body, new_env, new_store, new_time), seen, { case (config, seen) =>
          val Config(ae, env_, store, time) = config;  val new_time = config.tick
          val baddr = allocBind(x, new_time);  val new_env = env + (x -> baddr)
          val new_store = store.update(baddr, atomicEval(ae, env_, store))
          aeval(Config(e, new_env, new_store, new_time), seen, continue) }) })
    case ae if isAtomic(ae) => continue(config, new_seen) } }
\end{lstlisting}

The circular meta-continuation is abstracted by the @nd@ operator, and only one continuation
remains at the level of @aeval@. 
When encountering a set of possible closures, we apply @nd@ on the set with the latest explored
configurations and an anonymous function that works on each closure. The beginning part 
(before calling @aeval@) of this anonymous function is essentially abstracted from @makeMCont@, 
which constructs new environment and store for the closure being evaluated.
Then we apply @aeval@ on the body expression of the closure with a continuation, 
which is inlined from @new_cont@ and will invoke @aeval@ on @e@ with the current continuation of the @let@ expression.

As we can see, by using the @nd@ operator, the evaluation function can be simplified to a 
single-layer continuation-passing style.
The only continuation of @aeval@ plays the same role as the one in a concrete CPS interpreter.

\subsubsection{Values}

Up to now, the abstract interpreter still collects and returns a set of configurations.
But our target, the abstract definitional interpreters, should return a set of final \textit{values}
instead of collected configurations; the intermediate values are simply discarded.
To properly represent final values, we introduce the case class @VS@ which
contains a set of storable values, a timestamp, and a store.

@@

\subsubsection{Cache-Passing Style}

Recall that the latent assumption of the work-list algorithm in classical AAM is that
if we have seen a state @s@, it means we also have seen all the successors of state @s@.
The caching algorithm we adopt here replays the same assumption, but in a big-step
manner: if we have seen the configuration @c@, then it means we also have seen the
values that are evaluated from the configuration @c@.
Here we will apply the fixed-point caching algorithm as described from
\citeauthor{darais2017abstracting}'s ADI paper \cite{darais2017abstracting}.
The case class @Cache@ and its operations are defined as follow:

@@

We can now transform our abstract interpreter into cache-passing style. We begin by redefining the
return type @Ans@ as a case class which contains a @VS@ set and a cache. @Ans@ also implements
its own accumulating operation @++@ which defines as a union of @Set[VS]@s and a join of caches.
The join operation of caches is defined as joining its two store components.

@@

Accordingly, the @nd@ operator is slightly changed to cache-passing style. 
The argument of type @Ans@ in the continuation @k@ is replaced by an argument of type @Cache@;
every time when we apply @k@, the latest cache @acc.cache@ is provided.

As previously mentioned, the components of configurations are lifted as arguments to
@aeval@. Given the cache and the simplified one-layer continuation, the shape of function 
@aeval@ now looks as follows:

\begin{lstlisting}
  def aeval(e: Expr, env: Env, store: BStore, time: Time, cache: Cache, continue: Cont): Ans = {
    val config = Config(e, env, store, time);  val new_time = config.tick
    if (cache.outContains(config)) continue(Ans(cache.outGet(config), cache))
    else {
      val new_cache = cache.outUpdate(config, cache.inGet(config))
      e match { ... } } }
\end{lstlisting}

@@

@@

Now, we have a big-step abstract interpreter written in plain continuation-passing style which
utilizes the @nd@ operator and guarantees termination by caching.
